Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-10-03 experiments/2_supplement/supplement.Rmd
changes to supplement

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB).

@@

(ref:max-tol2) Relationship of ecosystem state variables and oxygen diffusivity when there is variation among nine strains of SB and the amount of variation is high. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol3) Relationship of ecosystem state variables and oxygen diffusivity when SB contained only the most tolerant strain, with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the ecosystem shown in Figure \@ref(fig:max-tol3), where the SB contained one strain with quite high tolerance, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -4.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:SB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation (Figure \@ref(fig:sb-div-effect4)). It appears that the high tolerance SB strain is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic state. This dynamical situation was not presented in the ecosystem in which high tolerance and low tolerance strains were present.

@@

## High diversity vs. high tolerance in CB and SB

@@

(ref:max-tol4)  Stable states found with nine strains of CB and SB and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol5)  Stable states found with the single most tolerant CB strain and SB strain and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the ecosystem shown in Figure \@ref(fig:max-tol5), where there was quite high tolerance CB strain and SB strain, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -6.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:CBSB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation. It appears that the high tolerance SB strain is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic stable state.

@@

## CB-SB-PB, high diversity

In the CB-SB-PB case when only the most tolerant strain of all functional groups is present, the resilience of the oxic state is greater than when variation in strain tolerance is present. Looked at from another perspective, comparing CB-SB and CB-SB-PB both with only the most tolerant strains, we see that presence of a high tolerance PB strain (in CB-SB-PB) allows the formation of an anoxic cyanobacteria-free stable state (whereas there was no such state in CB-SB). We interpret this as presence of a high tolerance PB strain stabilising the anoxic state, perhaps by increasing the availability of oxidised sulfur which is a resource for SB.

@@

(ref:max-tol7) Stable states found with the single most tolerant CB strain, SB strain, and PB strain, and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

### Why only at high diversity?

Why do these effects only occur at high diversities? The likely reason is that at high diversities then the most tolerant strain has particularly low growth rate, and so it may have the tolerance required to increase and persist, it does not have the growth rate required to dominate.

>>>>

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB). We show below that

@@

(ref:max-tol2) Stable states found when there is variation among nine strains of SB and the amount of variation is high. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol3) Stable states found when SB contained only the most tolerant strain, with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the ecosystem shown in Figure \@ref(fig:max-tol3), where the SB contained one strain with very high tolerance, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -4.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:SB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation (Figure \@ref(fig:sb-div-effect4)). It appears that the high tolerance SB strain is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic state. This dynamical situation was not present in the ecosystem in which high tolerance and low tolerance strains were present.

@@

## High tolerance vs. high diversity in CB and SB

@@

(ref:max-tol4)  Stables states found when there is high trait variation among nine strains of CB and SB. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol5)  Stable states found when CB and SB each contain only the single most tolerant strain with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the ecosystem shown in Figure \@ref(fig:max-tol5), where CB and SB each contained only one strain with very high tolerance, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -6.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:CBSB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation (Figure \@ref(fig:sb-div-effect4)). It appears that the SB strain with high tolerance is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic state.

@@

## High tolerance vs. high diversity in CB-SB-PB

When in each of the three functional groups only the most tolerant strain is present, the resilience of the oxic state is greater than when there is variation in strain tolerance within groups. Looked at from another perspective, comparing CB-SB and CB-SB-PB both with only the most tolerant strains, presence of a high tolerance PB strain (in CB-SB-PB) allows the formation of an anoxic cyanobacteria-free stable state (whereas there was no such state in CB-SB). We interpret this as presence of a high tolerance PB strain stabilising the anoxic state, perhaps by increasing the availability of oxidised sulfur which is a resource for SB. Nevertheless, absence of SB strains with high growth rate seems to shift the oxic-anoxic tipping point to lower oxygen diffusivity.

@@

(ref:max-tol7) Stable states found with the single most tolerant CB strain, SB strain, and PB strain, with trait values corresponding with the diversity level in the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

## Why only at high diversity/tolerance?

Why do these effects only occur at high diversity/tolerance? The likely reason is that at high diversities the most tolerant strain has particularly low growth rate, and so it may have the tolerance required to increase and persist, but not the growth rate required to dominate.
